NeuroElectro: A Window to the World's Neuron Electrophysiology Data

### 2014-03-17 neuroelectro_frontiers/main.tex
made changes to main.tex to respond to reviewer comments

\begin{table}[!t]
\textbf{\refstepcounter{table}\label{tab:journal_listing} Table \arabic{table}.}
{Statistics of journals represented in the NeuroElectro database. Listing of journals and counts of articles downloaded (articles obtained), articles with published data tables containing neurophysiological information which has been manually validated by an expert curator (validated), and articles which likely contain information in a data table which has not yet been manually curated (not validated). 
Not validated articles are those which have at least 4 algorithmically assigned electrophysiological concepts within a data table.}
\processtable{}
{\begin{tabular}{rrrr}
    \toprule
    Journal & Articles obtained & Validated & Not validated \\
    \midrule
    J. Neurosci. & 19002 & 104   & 560 \\
    J. Neurophysiol. & 12078 & 94    & 555 \\
    J. Physiol. (Lond.) & 10543 & 44    & 235 \\
    Neuroscience & 3035  & 14    & 205 \\
    Eur. J. Neurosci. & 2495  & 7     & 117 \\
    Brain Res. & 3017  & 7     & 146 \\
    Neuron & 1657  & 4     & 43 \\
    Epilepsia & 463   & 2     & 23 \\
    Neurosci. Lett. & 1468  & 2     & 34 \\
    Hippocampus & 208   & 2     & 10 \\
    \botrule
    \end{tabular}}[]
\end{table}

We obtained tens of thousands of potentially relevant full article texts directly from publisher websites.  
We first identified potential articles that were likely to contain information relevant to neuron biophysics using the native search functions provided within the journal websites and only downloaded articles containing a specific list of terms including ``input resistance'' and ``resting membrane potential'' (Fig. \ref{text_mining_workflow}).  
This pre-selection step allowed us to identify and download only articles that contained data relevant to our project.  
Upon identifying candidate articles, we then downloaded the full text of each potentially-relevant article as HTML; 
articles downloaded from the publisher Elsevier (e.g., \emph{Neuron and Brain Research}) were downloaded as XML using the provided text-mining API and subsequently converted to HTML.  
We chose to work with HTML (as opposed to PDF or XML) because HTML provides a machine-readable markup of the article's content, allowing us to easily identify relevant elements within the article -- such as data tables and the methods section -- using publicly available HTML-parsing tools (here we used the Beautiful Soup HTML-processing library implemented in Python: \url{http://www.crummy.com/software/BeautifulSoup/bs4/doc/}).  
Furthermore, because HTML is a single semi-structured standard used across publishers, we could write relatively generic HTML-processing algorithms applicable to content published across journals.  
Our focus on using HTML limits us to relatively newer articles - typically those published after 1996 - because before this time most publications are only available as scanned PDF files.  
However, because the rate of publication across the field has grown exponentially, this HTML-available subset constitutes the vast majority of neuroscience articles ever published.

@@

Given the challenges in mining raw electrophysiological traces from figure images, we instead focused on obtaining information about basic neuronal electrophysiological properties, such as input resistances and resting membrane potentials.  
Though this information is often presented within the text of the article, it is usually presented in complex sentence structures that are difficult to accurately parse algorithmically.  
Published data tables, on the other hand, present a unique opportunity for electrophysiological data extraction, since common techniques exist for extracting information from structured tables \citep{yarkoni_large-scale_2011}.  
Moreover, because tables succinctly summarize multiple attributes of a collected dataset, the effort of an expert curator can be put to best use when validating tables relative to validating content mined from article sentences or figure panels.  
While we estimate that only 5-10\% of electrophysiology articles contain data tables, there is sufficient redundancy within the field (i.e., multiple investigators often publish articles on the same neuron type) that focusing on data tables nevertheless yields substantial coverage of electrophysiological properties across nearly all major neuron types.

@@

When manually curating the text-mined content for the most commonly reported electrophysiological properties (resting membrane potential, input resistance, membrane time constant, spike half-width, spike amplitude, and spike threshold), we took care to account for and remove cases where the investigator had calculated an electrophysiological measurement using an inconsistent methodology (e.g., protocol or calculation differences).  
However we note that we could not identify all of these cases (in particular: spike amplitude, input resistance, and membrane time constant), in part because investigators did not always explicitly define how these measurements were calculated within their article.  
We note that in cases where we pool measurements which are measured using inconsistent protocols or calculations, this will tend to add unexplained variance to our data set.  
Given these measurement inconsistencies, we provide our recommendations for how these electrophysiological properties should be reported in future investigations via our electrophysiology ontology (see Supplemental Materials).

@@

\begin{figure*}[htbp!]
\centering
\includegraphics[scale = 1]{figures/metadata_assignment_accuracy_v2.jpg}
\caption{\textit{Accuracy of metadata assignment using automated methods alone.} }
\label{metadata_robot_accuracy_robot}
\end{figure*}

@@

\subsection{Scalability of current approach}
We note that multiple steps in our approach require manual intervention by an expert curator in order for electrophysiological measurements to be extracted with an acceptably low error rate.  
Namely, an expert curator needs to specify which neuron types are recorded from in each article and where data from the normotypic or ``control'' states of these neurons are textually referenced within a data table.  
Moreover, given the current accuracy of the unsurpervised algorithmic assignment of electrophysiological concepts and experimental metadata (78\% and 50\%, respectively), these also need to be manually validated and corrected as required by an expert.  
Given the necessity of these manual steps, the scalability of our current approach is limited by our ability to manually curate this information or by our ability to improve the error rate of the automated methods.  
Despite this limitation, our current pipeline is still much faster than a purely manual one.  The methodology could be further improved by addressing falsely matching entities (such as EPSP amplitude in section \ref{sec:ephys_extraction}) by either auto-suggesting alternative matches returned by string-similarity algorithms or by simply adding these concepts to the electrophysiolgical ontology.  
Moreover, these improvements would facilitate formally computing the sensitivity and specificity of these entity recognition methods.

@@

To facilitate further data integration efforts, we advocate small shifts in how investigators analyze and report neurophysiological data.  
Specifically, we recommend the usage of common standards and definitions for basic physiological measurements\citep{toledo-rodriguez_correlation_2004} and neuron types\citep{ascoli_petilla_2008,larson_neurolex.org:_2013}.  
We also ask that physiologists report more basic physiological information within articles and, if possible, publish such data using machine-readable formats like data tables.  
Similarly, relevant experimental details should be clearly stated within the methods section (e.g., liquid junction potential correction).  
These recommendations could be made informally by journals (in particular, requested by reviewers during manuscript review) as well as by funding agencies.

However, more widespread inclusion of fundamental electrophysiological measurements in data tables in future publications will strongly facilitate data-sharing and meta-analysis initiatives such as NeuroElectro.  
We believe that, if successful, the use of NeuroElectro will influence the practices of scientists writing papers and reporting results.  
Specifically we believe (and hope) that it will result in scientists reporting more basic physiological data overall as well as reporting more data in machine-parsable tables.  
This change would make it easier for scientists to find and make use of data collected by others.  
Moreover, such a culture shift has the potential to make science function more effectively and efficiently to facilitate discovery.

>>>>

\begin{table}[!t]
\textbf{\refstepcounter{table}\label{tab:journal_listing} Table \arabic{table}.}
{Statistics of journals represented in the NeuroElectro database. Listing of journals and counts of articles downloaded (articles obtained), articles with published data tables containing neurophysiological information which has been manually validated by an expert curator (validated), and articles which likely contain information in a data table which has not yet been manually curated (not validated). 
Not validated articles are those which have at least 4 algorithmically assigned electrophysiological concepts within a data table.}
\processtable{}
{\begin{tabular}{rrrr}
    \toprule
    Journal & Articles obtained & Validated & Not validated \\
    \midrule
    J. Neurosci. & 19002 & 104   & 560 \\
    J. Neurophysiol. & 12078 & 94    & 555 \\
    J. Physiol. (Lond.) & 10543 & 44    & 235 \\
    Neuroscience & 3035  & 14    & 205 \\
    Eur. J. Neurosci. & 2495  & 7     & 117 \\
    Brain Res. & 3017  & 7     & 146 \\
    Neuron & 1657  & 4     & 43 \\
    Epilepsia & 463   & 2     & 23 \\
    Neurosci. Lett. & 1468  & 2     & 34 \\
    Hippocampus & 208   & 2     & 10 \\
    \botrule
    \end{tabular}}
   {}
\end{table}

We obtained tens of thousands of potentially relevant full article texts directly from publisher websites.  
We first identified potential articles that were likely to contain information relevant to neuron biophysics using the native search functions provided within the journal websites and only downloaded articles containing a specific list of terms including ``input resistance'' and ``resting membrane potential'' (Fig. \ref{text_mining_workflow}).  
This pre-selection step allowed us to identify and download only articles that contained data relevant to our project.  
Upon identifying candidate articles, we then downloaded the full text of each potentially-relevant article as HTML; 
articles downloaded from the publisher Elsevier (e.g., \emph{Neuron and Brain Research}) were downloaded as XML using the provided text-mining API and subsequently converted to HTML.  
We chose to work with HTML (as opposed to PDF or XML) because HTML provides a machine-readable markup of the article's content, allowing us to easily identify relevant elements within the article -- such as data tables and the methods section -- using publicly available HTML-parsing tools (here we used the Beautiful Soup HTML-processing library implemented in Python: \url{http://www.crummy.com/software/BeautifulSoup/bs4/doc/}).  
Furthermore, because HTML is a single semi-structured standard used across publishers, we could write relatively generic HTML-processing algorithms applicable to content published across journals.  
Our focus on using HTML limits us to relatively newer articles - typically those published after 1996 - because before this time most publications are only available as scanned PDF files.  
However, because the rate of publication across the field has grown exponentially, this HTML-available subset constitutes the majority of published neuroscience articles.

@@

Given the challenges in mining raw electrophysiological traces from figure images, we instead focused on obtaining information about basic neuronal electrophysiological properties, such as input resistances and resting membrane potentials.  
Though this information is often presented within the text of the article, it is usually presented in complex sentence structures that are difficult to accurately parse algorithmically.  
Published data tables, on the other hand, present a unique opportunity for electrophysiological data extraction, since common techniques exist for extracting information from structured tables \citep{yarkoni_large-scale_2011}.  
Moreover, because tables succinctly summarize multiple attributes of a collected dataset, the effort of an expert curator can be put to best use when validating tables relative to validating content mined from article sentences or figure panels.  
While we estimate that only 5-10\% of electrophysiology articles contain data tables, there is sufficient redundancy within the field (i.e., multiple investigators often publish articles on the same neuron type) that focusing on data tables nevertheless yields substantial coverage of electrophysiological properties across many major neuron types.

@@

When manually curating the text-mined content for the most commonly reported electrophysiological properties (resting membrane potential, input resistance, membrane time constant, spike half-width, spike amplitude, and spike threshold), we took care to account for where the investigator had calculated an electrophysiological measurement using an inconsistent methodology (e.g., protocol or calculation differences) by normalizing such measurements to a common reference definition or removing such data when normalization was not possible.
However we note that we could not identify all of these cases (in particular: spike amplitude, input resistance, and membrane time constant), in part because investigators did not always explicitly define how these measurements were calculated within their article.  
We note that in cases where we pool measurements which are measured using inconsistent protocols or calculations, this will tend to add unexplained variance to our data set.  
Given these measurement inconsistencies, we provide our recommendations for how these electrophysiological properties should be reported in future investigations via our electrophysiology ontology (see Supplemental Materials).

@@

\begin{figure*}[htbp!]
\centering
\includegraphics[scale = .25]{figures/metadata_assignment_accuracy_v2.jpg}
\caption{\textit{Accuracy of metadata assignment using automated methods alone. Error indicate 95\% binomial confidence intervals.} }
\label{metadata_robot_accuracy_robot}
\end{figure*}

@@

\subsection{Scalability of current approach}
We note that multiple steps in our approach require manual intervention by an expert curator in order for electrophysiological measurements to be extracted with an acceptably low error rate.  
Namely, an expert curator needs to specify which neuron types are recorded from in each article and where data from the normotypic or ``control'' states of these neurons are textually referenced within a data table.  
Moreover, given the current accuracy of the unsurpervised algorithmic assignment of electrophysiological concepts and experimental metadata (78\% and 50\%, respectively), these also need to be manually validated and corrected and normalized as required by an expert.  
Given the necessity of these manual steps, the scalability of our current approach is limited by our ability to manually curate this information or by our ability to improve the error rate of the automated methods.  
Despite this limitation, our current pipeline is still much faster than a purely manual one.  The methodology could be further improved by addressing falsely matching entities (such as EPSP amplitude in section \ref{sec:ephys_extraction}) by either auto-suggesting alternative matches returned by string-similarity algorithms or by simply adding these concepts to the electrophysiolgical ontology.  
Moreover, these improvements would facilitate formally computing the sensitivity and specificity of these entity recognition methods.

@@

However, more widespread inclusion of fundamental electrophysiological measurements in data tables in future publications will strongly facilitate data-sharing and meta-analysis initiatives such as NeuroElectro.  
We believe that, if successful, the use of NeuroElectro will influence the practices of scientists writing papers and reporting results.  
Specifically, we recommend the usage of common standards and definitions for basic physiological measurements\citep{toledo-rodriguez_correlation_2004} and neuron types\citep{ascoli_petilla_2008,larson_neurolex.org:_2013}.  
Moreover, we advocate that, where possible, scientists report more basic physiological data overall and report such data using machine-parsable data tables.  
These recommendations could be made informally by journals (in particular, requested by reviewers during manuscript review) as well as by funding agencies.
This change would make it easier for scientists to find and make use of data collected by others.  
Moreover, such a culture shift has the potential to make science function more effectively and efficiently to facilitate discovery.
